# Chapter 1: Inference

## Normal Distribution

For these exercises, we will be using the following dataset:

```{r setup, include=TRUE}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
filename <- file.path("data", basename(url))
download(url, destfile=filename)
x <- unlist( read.csv(filename) )
write.csv(x, "data/femaleControlsPopulation.csv")
```

Here x represents the weights for the entire population.

1. What is the average of these weights?
```{r}
mean(x)
```

2. After setting the seed at 1, set.seed(1) take a random sample of size 5. What is the 
absolute value (use abs) of the difference between the average of the sample and the 
average of all the values?

```{r}
set.seed(1)
sample <- sample(x, 5)
abs(mean(sample) - mean(x))
```

3. After setting the seed at 5, set.seed(5) take a random sample of size 5. What is the 
absolute value of the difference between the average of the sample and the average of 
all the values?

```{r}
set.seed(5)
sample <- sample(x, 5)
abs(mean(sample) - mean(x))
```

4. Why are the answers from 2 and 3 different?  
**Answer:** C. Because the average of the samples is a random variable.

5. Set the seed at 1, then using a for-loop take a random sample of 5 mice 1,000 times. 
Save these averages. What percent of these 1,000 averages are more than 1 ounce away 
from the average of x ?

```{r}
set.seed(1)
n <- 1000
null <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 5)
    null[i] <- mean(sample)- mean(x)
}
mean(null > 1)
hist(null)
```

6. We are now going to increase the number of times we redo the sample from 1,000 to 10,000.
Set the seed at 1, then using a for-loop take a random sample of 5 mice 10,000 times. Save
these averages. What percent of these 10,000 averages are more than 1 ounce away from the
average of x ?

```{r}
set.seed(1)
n <- 10000
null <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 5)
    null[i] <- mean(sample)- mean(x)
}
mean(null > 1)
hist(null)
```

7. Note that the answers to 5 and 6 barely changed. This is expected. The way we think 
about the random value distributions is as the distribution of the list of values 
obtained if we repeated the experiment an infinite number of times. On a computer, we 
can’t perform an infinite number of iterations so instead, for our examples, we consider 1,000 to be large
enough, thus 10,000 is as well. Now if instead we change the sample size, then we change
the random variable and thus its distribution.

Set the seed at 1, then using a for-loop take a random sample of 50 mice 1,000 times. 
Save these averages. What percent of these 1,000 averages are more than 1 ounce away 
from the average of x ?

```{r}
set.seed(1)
n <- 1000
null <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 50)
    null[i] <- mean(sample)- mean(x)
}
mean(null > 1)

hist(null)
```

8. Use a histogram to “look” at the distribution of averages we get with a sample size of 5 and
a sample size of 50. How would you say they differ?  
**Answer:** B. They both look roughly normal, but with a sample size of 50 the spread is smaller.
When we have more samples, the difference between the mean of the sample and the mean of the 
population is smaller, i.e. we could infer the population mean better with more samples.

9. For the last set of averages, the ones obtained from a sample size of 50, what percent are
between 23 and 25?

```{r}
set.seed(1)
n <- 1000
avg <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 50)
    avg[i] <- mean(sample)
}
mean(avg > 23 & avg < 25)
hist(avg)
```


10. Now ask the same question of a normal distribution with average 23.9 and standard deviation 0.43.
`pnorm` computes the cumulative distribution function (CDF) of the normal distribution. 
It calculates the probability that a value is less than or equal to a given point.

```{r}
#calculating the probability that values are greater than 23 and 25 separately, rather than the probability that values are between 23 and 25.
# 1 - pnorm(q= c(23,25), mean = 23.9, sd= 0.43) 
#correct code
pnorm(q= 25, mean = 23.9, sd= 0.43) - pnorm(q= 23, mean = 23.9, sd= 0.43)
```

The answer to 9 and 10 were very similar. This is because we can approximate the distribution of
the sample average with a normal distribution. We will learn more about the reason for this next.

## Populations, Samples and Estimates

```{r}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- file.path("data", basename(url))
download(url, destfile=filename)
dat <- read.csv(filename)
dat <- na.omit(dat)
head(dat)
```

1. Use dplyr to create a vector x with the body weight of all males on the control (chow) diet.
What is this population’s average?

```{r}
library(dplyr)
m_chow <- dat %>%
    filter(Sex == "M", Diet == "chow") %>%
    select(Bodyweight) %>%
    unlist()
mean(m_chow)
```

2. Now use the rafalib package and use the popsd function to compute the population
standard deviation.

```{r}
#install.packages("rafalib")
library(rafalib)
popsd(m_chow)
```

3. Set the seed at 1. Take a random sample X of size 25 from x. What is the sample average?

```{r}
set.seed(1)
sample_m_chow <- sample(m_chow,25)
mean(sample_m_chow)
```

4. Use dplyr to create a vector y with the body weight of all males on the high fat (hf) diet.
What is this population’s average?

```{r}
m_hf <- dat %>%
    filter(Sex == "M", Diet == "hf") %>%
    select(Bodyweight) %>%
    unlist()
mean(m_hf)
```    

5. Now use the rafalib package and use the popsd function to compute the population
standard deviation.

```{r}
popsd(m_hf)
```

6. Set the seed at 1. Take a random sample Y of size 25 from y. What is the sample average?

```{r}
set.seed(1)
sample_m_hf <- sample(m_hf,25)
mean(sample_m_hf)
```

7. What is the difference in absolute value between mean(m_hf)-mean(m_chow) and mean(sample_m_chow)-mean(sample_m_hf)?

```{r}
#abs(abs((mean(y) - mean(x))) - abs((mean(sample_y) - mean(sample_x))))
abs((mean(m_hf) - mean(m_chow)) - (mean(sample_m_hf) - mean(sample_m_chow)))
```

8. Repeat the above for females. Make sure to set the seed to 1 before each sample call. What
is the difference in absolute value between mean(y)-mean(x) and mean(sample_y)-mean(sample_x)?

```{r}
#Population average and standard deviation for females mouse on chow diet
library(dplyr)
f_chow <- dat %>%
    filter(Sex == "F", Diet == "chow") %>%
    select(Bodyweight) %>%
    unlist()
mean(f_chow)
popsd(f_chow)

#Sample average for 25 females on chow diet
set.seed(1)
sample_f_chow <- sample(f_chow,25)
mean(sample_f_chow)
```

```{r}
#Population average and standard deviation for females mouse on hf diet
library(dplyr)
f_hf <- dat %>%
    filter(Sex == "F", Diet == "hf") %>%
    select(Bodyweight) %>%
    unlist()
mean(f_hf)
popsd(f_hf)

#Sample average for 25 females on hf diet
set.seed(1)
sample_f_hf <- sample(f_hf,25)
mean(sample_f_hf)
```

```{r}
abs((mean(f_hf) - mean(f_chow)) - (mean(sample_f_hf) - mean(sample_f_chow)))
```

9. For the females, our sample estimates were closer to the population difference than with
males. What is a possible explanation for this?  

**Answer:**  
- With seed=1, males closer (1.40 < 1.75). Females *typically* closer due to **lower population SDs**.  
- Typical correct answer: A. The population variance of the females is smaller than that of the males; thus, the
sample variable has less variability.

## Central Limit Theorem and t-distribution

### Central Limit Theorem
When the sample size is large, the average $\\bar{Y}$ of a random sample follows a normal distribution 
centered at the population average µ and with standard deviation equal to the population 
standard deviation σ, divided by the square root of the sample size N. We refer to the 
standard deviation of the distribution of a random variable as the random variable’s **standard error**.

```{r}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/mice_pheno.csv"
filename <- file.path("data", basename(url))
download(url, destfile=filename)
dat <- na.omit(read.csv(filename))
```

1. If a list of numbers has a distribution that is well approximated by the normal distribution,
what proportion of these numbers are within one standard deviation away from the list’s
average?

```{r}
pnorm(q=1, mean=0, sd=1) - pnorm(q=-1, mean=0, sd=1)
``` 
For the normal distribution: ~68% of values are within 1 SD

2. What proportion of these numbers are within two standard deviations away from the list’s
average?

```{r}
pnorm(q=2, mean=0, sd=1) - pnorm(q=-2, mean=0, sd=1)
``` 
For the normal distribution: ~95% of values are within 2 SDs

3. What proportion of these numbers are within three standard deviations away from the list’s
average?

```{r}
pnorm(q=3, mean=0, sd=1) - pnorm(q=-3, mean=0, sd=1)
``` 
For the normal distribution: ~99.7% of values are within 3 SDs

4. Define y to be the weights of males on the control diet. What proportion of the mice are
within one standard deviation away from the average weight (remember to use popsd for
the population sd)?

```{r}
y <- dat %>%
    filter(Sex == "M", Diet == "chow") %>%
    select(Bodyweight) %>%
    unlist()
mean(y)
popsd(y)
hist(y)

pnorm(q= mean(y)+ popsd(y), mean=mean(y), sd= popsd(y)) - pnorm(q= mean(y)- popsd(y), mean=mean(y), sd= popsd(y))
```
For y that approximately follows a normal distribution: ~68% of values are within 1 SD

5. What proportion of these numbers are within two standard deviations away from the list’s
average?

```{r}
pnorm(q= mean(y)+ 2*popsd(y), mean=mean(y), sd= popsd(y)) - pnorm(q= mean(y)- 2*popsd(y), mean=mean(y), sd= popsd(y))
```
For y that approximately follows a normal distribution: ~95% of values are within 2 SDs

6. What proportion of these numbers are within three standard deviations away from the list’s
average?
```{r}
pnorm(q= mean(y)+ 3*popsd(y), mean=mean(y), sd= popsd(y)) - pnorm(q= mean(y)- 3*popsd(y), mean=mean(y), sd= popsd(y))
```
For y that approximately follows a normal distribution: ~99.7% of values are within 3 SDs

7. Note that the numbers for the normal distribution and our weights are relatively close. Also,
notice that we are indirectly comparing quantiles of the normal distribution to quantiles
of the mouse weight distribution. We can actually compare all quantiles using a qqplot.

```{r}
qqnorm(y)
qqline(y)
```

**Answer:** C) The mouse weights are well approximated by the normal distribution, although
the larger values (right tail) are larger than predicted by the normal. This is consistent
with the differences seen between question 3 and 6. 

8. Create the above qq-plot for the four populations: male/females on each of the two diets. What
is the best explanation for all these being well approximated by the normal distribution?

```{r}
mypar(2,2)
#Male chow
y_m_chow <- dat %>%
    filter(Sex == "M", Diet == "chow") %>%
    select(Bodyweight) %>%
    unlist()
qqnorm(y_m_chow)
qqline(y_m_chow)        

#Male hf
y_m_hf <- dat %>%
    filter(Sex == "M", Diet == "hf") %>%
    select(Bodyweight) %>%
    unlist()
qqnorm(y_m_hf)
qqline(y_m_hf)      

#Female chow
y_f_chow <- dat %>%
    filter(Sex == "F", Diet == "chow") %>%
    select(Bodyweight) %>%
    unlist()
qqnorm(y_f_chow)
qqline(y_f_chow)    

#Female hf
y_f_hf <- dat %>%
    filter(Sex == "F", Diet == "hf") %>%
    select(Bodyweight) %>%
    unlist()
qqnorm(y_f_hf)
qqline(y_f_hf)
```

**Answer:** A) The CLT tells us that sample averages are approximately normal.

9. Here we are going to use the function `replicate` to learn about the distribution of random
variables. All the above exercises relate to the normal distribution as an approximation of the
distribution of a fixed list of numbers or a population. We have not yet discussed probability
in these exercises. If the distribution of a list of numbers is approximately normal, then if
we pick a number at random from this distribution, it will follow a normal distribution.
However, it is important to remember that stating that some quantity has a distribution does
not necessarily imply this quantity is random. Also, keep in mind that this is not related to
the central limit theorem. The central limit applies to averages of random variables. Let’s
explore this concept.

We will now take a sample of size 25 from the population of males on the chow diet. The
average of this sample is our random variable. We will use the replicate to observe 10,000
realizations of this random variable. Set the seed at 1, generate these 10,000 averages. Make
a histogram and qq-plot of these 10,000 numbers against the normal distribution.
We can see that, as predicted by the CLT, the distribution of the random variable is very
well approximated by the normal distribution.


```{r }
set.seed(1)
y <- filter(dat, Sex=="M" & Diet=="chow") %>% select(Bodyweight) %>% unlist
avgs <- replicate(10000, mean(sample(y, 25)))
mypar(1,2)
hist(avgs)
qqnorm(avgs)
qqline(avgs)
```

What is the average of the distribution of the sample average?

```{r}
mean(avgs)
```

10. What is the standard deviation of the distribution of sample averages?

```{r}
sd(avgs)
```

11. According to the CLT, the answer to exercise 9 should be the same as mean(y). You should
be able to confirm that these two numbers are very close. Which of the following does the
CLT tell us should be close to your answer to exercise 10?

```{r}
popsd(y)/sqrt(25)
```

12. In practice we do not know σ (popsd(y)) which is why we can’t use the CLT directly. This
is because we see a sample and not the entire distribution. We also can’t use popsd(avgs)
because to construct averages, we have to take 10,000 samples and this is never practical. We
usually just get one sample. Instead we have to estimate popsd(y). As described, what we
use is the sample standard deviation. Set the seed at 1, using the replicate function, create
10,000 samples of 25 and now, instead of the sample average, keep the standard deviation.
Look at the distribution of the sample standard deviations. It is a random variable. The real
population SD is about 4.5. What proportion of the sample SDs are below 3.5?

```{r}
set.seed(1)
y <- dat %>% filter(Sex=="M" & Diet=="chow") %>% select(Bodyweight) %>% unlist
sds <- replicate(10000, sd(sample(y, 25)))
mean(sds < 3.5)
```

13. What the answer to question 12 reveals is that the denominator of the t-test is a random
variable. By decreasing the sample size, you can see how this variability can increase.
It therefore adds variability. The smaller the sample size, the more variability is added.
The normal distribution stops providing a useful approximation. When the distribution of
the population values is approximately normal, as it is for the weights, the t-distribution
provides a better approximation. We will see this later on. Here we will look at the difference
between the t-distribution and normal. Use the function qt and qnorm to get the quantiles of
x=seq(0.0001,0.9999,len=300). Do this for degrees of freedom 3, 10, 30, and 100. Which
of the following is true?

```{r}
x <- seq(0.0001,0.9999,len=300)
mypar(3,2)
plot(x, qt(x, df=3), type="l", main="df=3")
plot(x, qt(x, df=10), type="l", main="df=10")
plot(x, qt(x, df=30), type="l", main="df=30")
plot(x, qt(x, df=100), type="l", main="df=100")
plot(x, qnorm(x), type="l", main="Normal")
lines(x, qnorm(x), col="red")
```

**Answer:**  C) The t-distribution has larger tails up until 30 degrees of freedom, at which point it
is practically the same as the normal distribution.

