# Chapter 1: Inference

For these exercises, we will be using the following dataset:

```{r setup, include=TRUE}
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleControlsPopulation.csv"
filename <- basename(url)
download(url, destfile=filename)
x <- unlist( read.csv(filename) )
write.csv(x, "data/femaleControlsPopulation.csv")
```

Here x represents the weights for the entire population.

1. What is the average of these weights?
```{r}
mean(x)
```

2. After setting the seed at 1, set.seed(1) take a random sample of size 5. What is the 
absolute value (use abs) of the difference between the average of the sample and the 
average of all the values?

```{r}
set.seed(1)
sample <- sample(x, 5)
abs(mean(sample) - mean(x))
```

3. After setting the seed at 5, set.seed(5) take a random sample of size 5. What is the 
absolute value of the difference between the average of the sample and the average of 
all the values?

```{r}
set.seed(5)
sample <- sample(x, 5)
abs(mean(sample) - mean(x))
```

4. Why are the answers from 2 and 3 different?  
**Answer:** C. Because the average of the samples is a random variable.

5. Set the seed at 1, then using a for-loop take a random sample of 5 mice 1,000 times. 
Save these averages. What percent of these 1,000 averages are more than 1 ounce away 
from the average of x ?

```{r}
set.seed(1)
n <- 1000
null <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 5)
    null[i] <- mean(sample)- mean(x)
}
mean(null > 1)
hist(null)
```

6. We are now going to increase the number of times we redo the sample from 1,000 to 10,000.
Set the seed at 1, then using a for-loop take a random sample of 5 mice 10,000 times. Save
these averages. What percent of these 10,000 averages are more than 1 ounce away from the
average of x ?

```{r}
set.seed(1)
n <- 10000
null <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 5)
    null[i] <- mean(sample)- mean(x)
}
mean(null > 1)
hist(null)
```

7. Note that the answers to 5 and 6 barely changed. This is expected. The way we think 
about the random value distributions is as the distribution of the list of values 
obtained if we repeated the experiment an infinite number of times. On a computer, we 
can’t perform an infinite number of iterations so instead, for our examples, we consider 1,000 to be large
enough, thus 10,000 is as well. Now if instead we change the sample size, then we change
the random variable and thus its distribution.

Set the seed at 1, then using a for-loop take a random sample of 50 mice 1,000 times. 
Save these averages. What percent of these 1,000 averages are more than 1 ounce away 
from the average of x ?

```{r}
set.seed(1)
n <- 1000
null <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 50)
    null[i] <- mean(sample)- mean(x)
}
mean(null > 1)

hist(null)
```

8. Use a histogram to “look” at the distribution of averages we get with a sample size of 5 and
a sample size of 50. How would you say they differ?  
**Answer:** B. They both look roughly normal, but with a sample size of 50 the spread is smaller.
When we have more samples, the difference between the mean of the sample and the mean of the 
population is smaller, i.e. we could infer the population mean better with more samples.

9. For the last set of averages, the ones obtained from a sample size of 50, what percent are
between 23 and 25?

```{r}
set.seed(1)
n <- 1000
avg <- vector("numeric",n)
for (i in 1:n) {
    sample <- sample(x, 50)
    avg[i] <- mean(sample)
}
mean(avg > 23 & avg < 25)
hist(avg)
```


10. Now ask the same question of a normal distribution with average 23.9 and standard deviation 0.43.
`pnorm` computes the cumulative distribution function (CDF) of the normal distribution. 
It calculates the probability that a value is less than or equal to a given point.

```{r}
#calculating the probability that values are greater than 23 and 25 separately, rather than the probability that values are between 23 and 25.
# 1 - pnorm(q= c(23,25), mean = 23.9, sd= 0.43) 
#correct code
pnorm(q= 25, mean = 23.9, sd= 0.43) - pnorm(q= 23, mean = 23.9, sd= 0.43)
```

The answer to 9 and 10 were very similar. This is because we can approximate the distribution of
the sample average with a normal distribution. We will learn more about the reason for this next.